---
title: "wiod.diagrammer -- R package for an easy work with WIOD (the 2016 release) including diagramming (flowcharts)"
author: "Aleksander Rutkowski"
date: "`r format(Sys.Date())`"
output: github_document
always_allow_html: yes
---

```{r setup, include=FALSE}
# Rendering this doc:
# rmarkdown::render(paste0(getwd(),'/README.Rmd'), envir=globalenv())
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(engine = 'R')
```

## Installation

```{r, eval=FALSE}
## if package `devtools` not installed, first do this:
# install.packages('devtools')
devtools::install_github('alekrutkowski/wiod.diagrammer')
```

## Documentation

https://alekrutkowski.github.io/wiod.diagrammer/reference/index.html

## Example

Load WIOD data for a given year from an official WIOD .Rdata file
(release 2016), which has to be first downloaded from
http://www.wiod.org/protected3/data16/wiot_ROW/wiot_r_Nov16.zip
and extracted (unzipped).

```{r library, warning=FALSE, message=FALSE}
library(wiod.diagrammer)
```

```{r rmd_helpers, include=FALSE}
library(magrittr)
plotLinks <- function(...) # a trick to plot svg
    wiod.diagrammer::plotLinks(...) %>% 
    DiagrammeRsvg::export_svg() %>% 
    cat(file='Graph.svg')
print <- function(d) # Just for clarity
    `if`(d %>% is.data.frame,
         base::print.data.frame(`colnames<-`(d, paste0(colnames(d),'  ')),
                                right=FALSE,
                                row.names=FALSE),
         base::print.default(d))
```

```{r load, cache=TRUE}
W <- loadWIOD('WIOT2014_October16_ROW.RData')
```

```{r  exploreW}
# Check the names of the first 10 columns:
cat(head(colnames(W), 20),
    sep='\n')
# How many columns and rows?
message(ncol(W),' columns; ',nrow(W),' rows')
```

Now let's flatten (reshape into long format):

```{r flattenW}
W_flat <- flatWIOD(W)
# See the structure of W:
str(W_flat)
head(W_flat, 10)
tail(W_flat, 10)
```

`wiod.diagrammer` uses internally the
[`data.table`](https://cran.r-project.org/web/packages/data.table/index.html)
package for performance. You can work with `W_flat` using the
`data.table`'s semantcs (e.g. the `:=` in-place column generation/modification)
or, if you prefer base R `data.frame`s, you may convert `W_flat` into
a regular `data.frame` with the function `as.data.frame`, do the modifications,
and then convert it back into `data.table` with the function
`data.table::as.data.table` for further processing (through the function
`findLinks` described below).

Now let's find the top supplier-user linkages in WIOD (as defined by the
column `value` in `W_flat` which -- if not modified -- corresponds to the
intermediate consumption in `W`).

First let's introduce a helper function `tieRobustRankLessOrEqual` (available
in `wiod.diagrammer`) comparing it with base R `rank`:

```{r tieRobustRankLessOrEqual}
x <- c(1,1,2,2,2,3,3)
# A comparison:
print(data.frame(x,
                 rank(x) <= 2,
                 tieRobustRankLessOrEqual(x, 2),
                 check.names = FALSE))
```

Now, let's find e.g. the 3 main customers of German automotive industry
as well as 2 main customers of those
customers and 1 main customer of those customer's customers (we could go
even deeper if we wanted or use different cut-off ranks, not
necessarily in the declining order).
So, we have 3 levels of linkages (the 1st one -- direct, the 2nd one and
the 3rd one -- indirect). NB: Some industries may re-emerge at different
rounds and they may be both customers and suppliers at the same time.
Let's also differentiate our ranks (top 3, 2, and 1) by
one more dimension: domestic vs foreign linkages.

```{r toplLinkagesCustomers}
W_flat[, domestic :=   # creating column in-place, following data.table's semantics
           ExpCountry == ImpCountry]
# Let's get rid of self-produced intermediate consumption
W_flat_noself <- W_flat[!(domestic &
                              ExpSectorNr == ImpSectorNr)]
# Let's keep only intra-EU trade
COUNTRIES_DT <- countries()
EU_COUNTRIES <-
    COUNTRIES_DT$Country[COUNTRIES_DT$isEUmember]
# Let's keep only flows >= 1 billion USD for clarity
W_flat_noself_truncated <- W_flat_noself[value >= 1000 &  # original WIOD data is in million USD
                                             ExpCountry %in% EU_COUNTRIES &
                                             ImpCountry %in% EU_COUNTRIES]
TOP_CUSTOMERS <-
    findLinks(partners = 'users',
              flat_wiod = W_flat_noself_truncated,
              start_countries = 'DEU', # We could add here other countries.
              start_sectors = 20, # "Manufacture of motor vehicles, trailers and semi-trailers"
              # We could add here other sectors.
              ListOfselectionFuns = # As discussed in the text above:
                  list(function(flat_wiod) tieRobustRankLessOrEqual(-flat_wiod$value, 3),  # minus because we
                       function(flat_wiod) tieRobustRankLessOrEqual(-flat_wiod$value, 2),  # want to rank from
                       function(flat_wiod) tieRobustRankLessOrEqual(-flat_wiod$value, 1)), # the highest to the lowest
              by = c('domestic','ExpCountry','ExpSectorNr')) # as discussed above
str(TOP_CUSTOMERS)
head(TOP_CUSTOMERS, 20)
```

Now let's do a similar exercise, just "upstream" i.e. for suppliers of suppliers.

```{r toplLinkagesSuppliers}
TOP_SUPPLIERS <-
    findLinks(partners = 'suppliers',
              flat_wiod = W_flat_noself_truncated,
              start_countries = 'DEU', # We could add here other countries.
              start_sectors = 20, # "Manufacture of motor vehicles, trailers and semi-trailers"
              # We could add here other sectors.
              ListOfselectionFuns = # As discussed in the text above:
                  list(function(flat_wiod) tieRobustRankLessOrEqual(-flat_wiod$value, 3),  # minus because we
                       function(flat_wiod) tieRobustRankLessOrEqual(-flat_wiod$value, 2),  # want to rank from
                       function(flat_wiod) tieRobustRankLessOrEqual(-flat_wiod$value, 1)), # the highest to the lowest
              by = c('domestic','ImpCountry','ImpSectorNr')) # as discussed above
str(TOP_SUPPLIERS)
head(TOP_SUPPLIERS, 20)
```

Now, let's plot the "upstream" linkages, making all the German sectors blue.
By default, the cross-border flows are dashed, while the domestic flows are
solid lines (arrows). The numbers in the nodes (rectangles) represent,
by default, the sector output (or the total intermediate consumption for the
final use sectors such as final consumption or investment if they show up
in the customers' graphs).

In `wiod.diagrammer` the rendering of the plot is done internally by [`DiagrammeR::grViz`](https://www.rdocumentation.org/packages/DiagrammeR/topics/grViz).
The plots can be saved manually in RStudio, or programmatically e.g. to an .svg file via
[`DiagrammeRsvg::export_svg`](https://www.rdocumentation.org/packages/DiagrammeRsvg/topics/export_svg)
to a character vector and then [`cat`ed](https://www.rdocumentation.org/packages/base/topics/cat),
or to a .png file piping them through
[`DiagrammeRsvg::export_svg`](https://www.rdocumentation.org/packages/DiagrammeRsvg/topics/export_svg),
[`charToRaw`](https://www.rdocumentation.org/packages/base/topics/charToRaw) and [`rsvg::rsvg_png`](https://www.rdocumentation.org/packages/rsvg/topics/rsvg_png).

```{r plot, message=FALSE, results='hide', warning=FALSE}
plotLinks(top_links_dt = TOP_SUPPLIERS,
		  wiot = W, # this is necessary
		  specificNodeOptionsFun =  # this is optional, just to show-off:
		  	function(country_sector_dt)
		  		ifelse(country_sector_dt$Country=='DEU',
		  			   'style=filled, fillcolor=cadetblue1', "")) # GraphViz colour names can be found at:
                                                                  # http://www.graphviz.org/doc/info/colors.html
```

Click on the picture to zoom in:

![Graph](https://cdn.rawgit.com/alekrutkowski/wiod.diagrammer/master/Graph.svg)

What else is available in the package? The functions which produce
auxiliary `data.table`s (that are used by `wiod.diagrammer`'s `plotLinks`
function if evaluated with default argument values).

```{r whatElse1}
COUNTRIES <- countries() # NB: no argument to function `countries`
str(COUNTRIES)
print(COUNTRIES)
SECTORS <- sectors(W)
SECTORS[, SectorLab :=  # truncate the long sector labels just for clarity below
            substr(SectorLab, 1, 30)]
str(SECTORS)
head(SECTORS, 20)
AGGREGATES <- aggregates(W)
```

The variable/column names produced by the function `aggregates`
reflect those in WIOD. They are explained in the documentations
of `aggregates`.

```{r whatElse2}
str(AGGREGATES) 
head(AGGREGATES, 20)
```


